The Journey from Wild to Text Book Data: A Case Study from the National Longitudinal Survey of Youth

### 2021-09-07 paper/paper.Rmd
Edit EDA part and add it's script.

read_chunk(here::here("paper/scripts/data-preprocessing.R"))
read_chunk(here::here("paper/scripts/ida.R"))
```

@@

```{r}
wages <- yowie::wages %>%
  mutate(year0 = year - 1979,
         ln_wages = log(wage))

wages_slope <- key_slope(wages, ln_wages ~ year0) %>%
  left_join(yowie::demog_nlsy79, by = "id") %>%
  mutate(hgc_regroup = fct_collapse(hgc, `ungraded` = "UNGRADED",
                                    `8TH grade or less` = c("1ST GRADE", "3RD GRADE",
                                                                 "4TH GRADE", "5TH GRADE",
                                                                 "6TH GRADE", "7TH GRADE",
                                                                 "8TH GRADE"),
                                    `9TH to 11TH grade` = c("9TH GRADE",
                                                            "10TH GRADE",
                                                            "11TH GRADE"),
                                    `12TH grade` = "12TH GRADE")) %>%
  mutate(hgc_regroup = factor(hgc_regroup, levels = c("UNGRADED",
                                                      "8TH grade or less", 
                                                      "9TH to 11TH grade", 
                                                      "12TH grade")))

```

This part gives some examples of exploratory data analysis using the data. In this example, we are interested to explore how people experience changes in their wages over the years. Hence, we focus the analysis in their slope of wages. Using `key_slope` function in `brolgar` [@brolgar], we perform linear regression with the mean hourly wage as response variable and year as the predictor for each individual.

Exploring the wages slope of each individual:
  - Exploring the wages slope pattern based on the number of observation

```{r slope-gender, fig.cap="slope-gender"}
slope_bottom_5<- wages_slope %>%
  arrange(.slope_year0) %>%
  slice_head(n = 5) %>%
  arrange(desc(.slope_year0)) %>%
  mutate(note = "bottom 5") %>%
  mutate(rank = seq_along(1:5))

slope_top_5<- wages_slope %>%
  arrange(desc(.slope_year0)) %>%
  slice_head(n = 5) %>%
  mutate(note = "top 5") %>%
  mutate(rank = seq_along(1:5))

keep_contrast <- rbind(slope_bottom_5, slope_top_5) %>%
  dplyr::select(.slope_year0, id, note, rank, hgc_regroup)

contrast_slope <- left_join(keep_contrast, wages, by = "id") %>%
  arrange(.slope_year0) %>%
  mutate(id = factor(id, levels = c("7918", "11695", "9176", "10119", "6464", "1271", 
                                    "12371", "9039", "10217", "7782")),
         note = factor(note, levels = c("top 5", "bottom 5")))

ggplot(contrast_slope, aes(x = year, y = ln_wages, color = gender)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_grid(note ~ rank) +
  scale_color_manual(values = c("#04B4AE", "#BF00FF"))

```

```{r slope-educ}
ggplot(contrast_slope, aes(x = year, y = ln_wages, color = hgc_regroup)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_grid(note ~ rank)
```

```{r slope-nobs}
wages_obs <- wages %>%
  features(wage, n_obs) %>%
  left_join(wages_slope, by = "id")

ggplot(wages_obs, aes(x = n_obs, y = .slope_year0)) +
  geom_point() +
  geom_jitter()
```

@@

```{r wages-gender-gap, echo=FALSE, fig.cap="Gender wage gap or the median wages of females relatives to median wages of males. Females tend to earn fewer than males. For example, in 2018 females earn about 78 cents for every $1 earned by males."}

wages_gender <- as.tibble(wages) %>%
  group_by(gender, year) %>%
  summarise(mean = mean(wage),
            median = median(wage),
            sd = sd(wage),
            samplesize = n()) %>%
  ungroup() %>%
  mutate(upper = mean + sd/sqrt(samplesize),
         lower = mean - sd/sqrt(samplesize))

wage_female <- wages_gender %>%
  filter(gender == "FEMALE") %>%
  dplyr::select(year, median) %>%
  rename(female = median)

wage_male <- wages_gender %>%
  filter(gender == "MALE") %>%
  dplyr::select(median) %>%
  rename(male = median)

gender_wage_gap <- cbind(wage_female, wage_male) %>%
  mutate(gap = (male-female)/male) %>%
  mutate(relative = 1-gap)

ggplot(gender_wage_gap) +
  geom_line(aes(x = year,
                y = relative),
            color = "#DF7401", size = 0.8) +
  geom_point(aes(x = year,
                y = relative),
             color = "#DF7401", size = 2) +
  geom_hline(yintercept = 1, colour = "#088A68", size = 1) +
  scale_x_continuous(limits = c(1978.7,2018.3), expand = c(0, 0)) +
  ylab("females' median wages relative to males (US$)") +
  annotate("text", x = 2014.5, y = 1, label = "$1 of male wages", vjust = 1.4) +
  annotate("text", x = 2012, y = 0.8, label = "female wages relative to male", vjust = 1)
```

>>>>

read_chunk(here::here("paper/scripts/data-preprocessing.R"))
read_chunk(here::here("paper/scripts/ida.R"))
read_chunk(here::here("paper/scripts/eda.R"))
```

@@

This part gives some examples of exploratory data analysis using the data. We are interested to explore how people experience changes in their wages over the years. Hence, we focus the analysis in their slope of wages. Using `key_slope` function in `brolgar` [@brolgar], we perform linear regression with the mean hourly wage as response variable and year as the predictor for each individual.

```{r nobs-plot, fig.cap="Distribution of observations. The numbers of rounds varies among respondents. Most of respondents participated in 28 rounds of survey."}
```

```{r slope}
```

```{r slope-nobs, fig.cap="Distribution of wages slope by number of observations. The slopes' range are more stable with more observations."}
```

```{r trim-slope}
```

```{r slope-gender, fig.cap = "Wages slope over the years based on gender. Most of respondents who experienced high wages slope are males. In contrast, respondents who experienced lower wages slope are mostly females."}
```

@@

```{r wages-gender-gap, echo=FALSE, fig.cap="Gender wage gap or the median wages of females relatives to median wages of males. Females tend to earn fewer than males. For example, in 2018 females earn about 78 cents for every $1 earned by males. Note that this is a crude measure as gap is calculated not based on other variable, such as experience and job class."}
```

```{r slope-educ, fig.cap = "Wages slope over the years based on highest education completed. People who are in top-5 and bottom-5 wages are dominated by people who completed high school."}
```

### 2021-09-07 paper/paper.tex
Edit EDA part and add it's script.

This part gives some examples of exploratory data analysis using the data. In this example, we are interested to explore how people experience changes in their wages over the years. Hence, we focus the analysis in their slope of wages. Using \texttt{key\_slope} function in \texttt{brolgar} (Tierney, Cook, and Prvan 2020), we perform linear regression with the mean hourly wage as response variable and year as the predictor for each individual.

Exploring the wages slope of each individual:
- Exploring the wages slope pattern based on the number of observation

@@

\caption{slope-gender}\label{fig:slope-gender}
\end{figure}

\begin{center}\includegraphics[width=\textwidth]{figures/slope-educ-1} \end{center}

\begin{center}\includegraphics[width=\textwidth]{figures/slope-nobs-1} \end{center}

@@

\textbackslash caption\{Gender wage gap or the median wages of females relatives to median wages of males. Females tend to earn fewer than males. For example, in 2018 females earn about 78 cents for every \$1 earned by males.\}\label{fig:wages-gender-gap}
\textbackslash end\{figure\}

>>>>

This part gives some examples of exploratory data analysis using the data. We are interested to explore how people experience changes in their wages over the years. Hence, we focus the analysis in their slope of wages. Using \texttt{key\_slope} function in \texttt{brolgar} (Tierney, Cook, and Prvan 2020), we perform linear regression with the mean hourly wage as response variable and year as the predictor for each individual.

{\centering \includegraphics[width=\textwidth]{figures/nobs-plot-1}

\caption{Distribution of observations. The numbers of rounds varies among respondents. Most of respondents participated in 28 rounds of survey.}\label{fig:nobs-plot}
\end{figure}

{\centering \includegraphics[width=\textwidth]{figures/slope-nobs-1}

\caption{Distribution of wages slope by number of observations. The slopes' range are more stable with more observations.}\label{fig:slope-nobs}
\end{figure}

@@

\caption{Wages slope over the years based on gender. Most of respondents who experienced high wages slope are males. In contrast, respondents who experienced lower wages slope are mostly females.}\label{fig:slope-gender}
\end{figure}

@@

\textbackslash caption\{Gender wage gap or the median wages of females relatives to median wages of males. Females tend to earn fewer than males. For example, in 2018 females earn about 78 cents for every \$1 earned by males. Note that this is a crude measure as gap is calculated not based on other variable, such as experience and job class.\}\label{fig:wages-gender-gap}
\textbackslash end\{figure\}

{\centering \includegraphics[width=\textwidth]{figures/slope-educ-1}

\caption{Wages slope over the years based on highest education completed. People who are in top-5 and bottom-5 wages are dominated by people who completed high school.}\label{fig:slope-educ}
\end{figure}
